---
title: "Host-parasitoid model description"
author: "Lucas A. Nell"
execute:
  echo: false
  message: false
  warning: false
format: 
  html:
    html-math-method: katex
vignette: >
  %\VignetteIndexEntry{Host-parasitoid model description}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


```{=html}
<style type="text/css">
  figure.quarto-float-tbl figcaption.quarto-float-caption-top {
    margin-top: .5rem;
    margin-bottom: .25rem;
    text-align:left
  }
</style>
```


```{r}
#| label: setup
#| include: false
library(knitr)
library(kableExtra)
library(gt)
library(gameofclones)
library(aeonia)
# Make a number print in LaTeX friendly format:
dbl4ltx <- function(x, .digits = 4) {
    .fmt <- sprintf("$%%.%if$", .digits)
    x <- sprintf(.fmt, x)
    return(I(x))
}
```




## Aphid population dynamics

See @tbl-aphids for parameters:
<!-- -->
\begin{equation}
\begin{split}
    \mathbf{X}(t+1) &= ( 1 - m ) \, ( 1 - B ) \, S(z(t)) \, 
        \mathbf{A}(x(t), Y(t)) \odot \left\{ \mathbf{L} \, \mathbf{X}(t) \right\} \\
    S(z(t)) &= \left( 1 + \frac{z(t)}{K} \right)^{-1} \\
    \mathbf{L} &= \begin{pmatrix}
        s_j     & f_N(z(t)) & 0     & f   \\
        r       & s_a       & 0     & 0   \\
        0       & f_W(z(t)) & s_j   & 0   \\
        0       & 0         & r     & s_a \\
        \end{pmatrix} \\
    f_N(z(t)) &= f \, \left[1 - p_W(z(t))\right] \\
    f_W(z(t)) &= f \, p_W(z(t)) \\
    p_W(z(t)) &= \textrm{logit}^{-1}(\alpha_0 + \alpha_z \, z(t)) \\
\end{split}
\end{equation}


```{r}
#| label: tbl-aphids
#| tbl-cap: "Model parameters for aphid population dynamics, including the value(s) used in simulations. For variables that change through time, the **Used** column is '$-$'."
tribble(
    ~Parameter,  ~Used,  ~Description ,
    "$n$"    ,  "$2$", "number of aphid age stages",
    "$\\mathbf{X}(t)$",  "$-$", "length-$2n$ vector of aphid abundances by stage",
    "$m$"    ,  dbl4ltx(pop_info$m, 1), "mortality from generalist predators and host-plant death",
    "$B$"    ,  "varies", "mortality from *Pseudomonas* bacteria (0 if not present on plant)",
    "$S(z(t))$",  "$-$",  "unparasitized aphid density dependence",
    "$z(t)$",  "$-$",  "total alive aphids in a particular patch",
    "$\\mathbf{A}$",  "$-$",  "length-$2n$ vector of the proportion of aphids surviving parasitism",
    "$x(t)$",  "$-$",  "total unparasitized aphids in a particular patch",
    "$Y(t)$",  "$-$",  "total adult, female parasitoids",
    "$\\mathbf{L}$",  "$**$",  "$2n \\times 2n$ transition matrix that defines transitions among all stages",
    "$K$",  dbl4ltx(pop_info$K, 0),  "controls strength of unparasitized aphid density dependence",
    "$s_j$",  dbl4ltx(pop_info$surv_j),  "aphid juvenile survival",
    "$s_a$",  dbl4ltx(pop_info$surv_a),  "aphid adult survival",
    "$r$",  dbl4ltx(pop_info$recruit),  "aphid recruitment",
    "$f$",  dbl4ltx(pop_info$fecund),  "aphid fecundity",
    "$f_N(z(t))$",  "$-$",  "non-winged aphid fecundity for non-winged offspring",
    "$f_W(z(t))$",  "$-$",  "non-winged aphid fecundity for winged offspring",
    "$p_W(z(t))$",  "$-$",  "proportion of winged offspring from non-winged mothers",
    "$\\alpha_0$",  dbl4ltx(pop_info$alate_0, 0),  "affects the proportion of winged offspring at low aphid densities",
    "$\\alpha_z$",  dbl4ltx(pop_info$alate_1),  "effect of aphid density on winged-aphid production",
) |> 
    kable(escape = FALSE)
```




## Parasitized aphids, mummies, adult parasitoids


See Table @tbl-parasitized for new parameters:
<!-- -->
\begin{equation}
\begin{split}
    \mathbf{A}(x(t), Y(t)) &=  \left[1 + \mathbf{R} \, \frac{a \, Y(t) }{
        k \, (h \, x(t) + 1)} \right]^{\circ -k} \\
    P(t+1) &= ( 1 - m ) \, ( 1 - B ) \, S_p(z(t)) \, \left(s_p \, P(t) + 
        \left[1 - \mathbf{A}(x(t),Y(t))\right]^{\intercal} \,
            \mathbf{L} \mathbf{X}(t)
        \right) - \tilde{\theta}_{p} P(t) \\
    S_p(z(t)) &= \left( 1 + \frac{z(t)}{K_p} \right)^{-1} \\
    M(t+1) &= ( 1 - m ) \, \left[M(t) + \tilde{\theta}_{p} \, P(t) \right] - \tilde{\theta}_{m} \, M(t) \\
    Y(t+1) &= s_y \, Y(t) + \frac{1}{2} \, \tilde{\theta}_{m} \, M(t) \\
\end{split}
\end{equation}

In this case, the following must be true to avoid negative numbers:
<!-- -->
\begin{equation}
\begin{split}
    \tilde{\theta}_{m} &< 1 - m \\
    \tilde{\theta}_{p} &< (1 - m) \, (1 - B) \, S_p(z(t)) \, s_p \\
\end{split}
\textrm{,}
\end{equation}
<!-- -->
\noindent where the minimum value of $S_p(z(t))$ is $\sim 0.65$ 
even when non-parasitized aphids are at their carrying capacity
($K \, (\lambda - 1)$, where $\lambda$ is the leading 
eigenvalue of $\mathbf{L}$),
which is higher than for parasitized aphids.


```{r}
#| label: tbl-parasitized
#| tbl-cap: "Model parameters for parasitized aphids, mummies, and adult parasitoids, including the value(s) used in simulations. For variables that change through time, the **Used** column is '$-$'. ** indicates that the vector used for $\\mathbf{R}$ was $\\\\{0.1861, 0.0053, 0.1861, 0\\\\}$"
tribble(
    ~Parameter,  ~Used,  ~Description ,
    "$\\mathbf{R}$", "$**$", "length-$2n$ vector giving the relative parasitoid attack rates by aphid stage",
    "$a$"    , dbl4ltx(pop_info$a, 2), "overall parasitoid attack rate",
    "$h$"    , dbl4ltx(pop_info$h, 3), "parasitoid attack-rate handling time",
    "$k$"    , dbl4ltx(pop_info$k, 2), "negative binomial distribution's aggregation parameter",
    "$P$"    , "$-$", "parasitized aphid abundance",
    "$S_p(z(t))$"  , "$-$", "parasitized aphid density dependence",
    "$K_p$"  , dbl4ltx(pop_info$K_p, 0), "controls strength of parasitized aphid density dependence",
    "$s_p$"  , dbl4ltx(pop_info$s_p), "daily survival for parasitized aphids",
    "$\\tilde{\\theta}_{p}$", dbl4ltx(pop_info$theta_p), "proportion of parasitized aphids that transition to mummies each day",
    "$M$"  , "$-$", "mummy (parasitized, dead aphid) abundance",
    "$\\tilde{\\theta}_{m}$", dbl4ltx(pop_info$theta_m), "proportion of mummies that transition to adult parasitoids each day",
    "$s_y$"  , dbl4ltx(pop_info$s_y, 2), "adult parasitoid daily survival",
) |> 
    kable(escape = FALSE)
```
