---
title: "Host-parasitoid model description"
author: "Lucas A. Nell"
execute:
  echo: false
  message: false
  warning: false
format: 
  html:
    html-math-method: katex
vignette: >
  %\VignetteIndexEntry{Host-parasitoid model description}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


```{=html}
<style type="text/css">
  figure.quarto-float-tbl figcaption.quarto-float-caption-top {
    margin-top: .5rem;
    margin-bottom: .25rem;
    text-align:left
  }
</style>
```


```{r}
#| label: setup
#| include: false
library(knitr)
library(kableExtra)
library(gt)
library(gameofclones)
library(aeonia)
# Make a number print in LaTeX friendly format:
dbl4ltx <- function(x, .digits = 4) {
    .fmt <- sprintf("$%%.%if$", .digits)
    x <- sprintf(.fmt, x)
    return(I(x))
}
```


In the equations below, $\odot~$indicates element-wise
multiplication (Hadamard product), 
$^{\circ}~$indicates element-wise power (Hadamard power),
and $^{\intercal}~$indicates matrix transposition.
Vectors and matrices are in bold.
Unless otherwise noted, indexing occurs first by space, then by time,
so $x_{it}$ is for patch $i$ at time $t$.


## Aphid population dynamics

Aphid populations ($\mathbf{X}$) on patch $i$ follow the following equations
(see @tbl-aphids for parameters):


<!-- -->
$$
\begin{split}
    \mathbf{X}_{i,t+1} &= g \, \psi_i \, S(z_{it}) \, 
        \mathbf{A}(x_{it}, \tilde{z}_{it}, Y_{t}) \odot \left[ \mathbf{L} \, \mathbf{X}_{it} \right] \\
    S(z_{it}) &= \left( 1 + \frac{z_{it}}{K} \right)^{-1} \\
    \mathbf{L} &= \begin{bmatrix}
        u     & f \, (1{-}W(z_{it})) & 0     & f   \\
        r       & s       & 0     & 0   \\
        0       & f \, W(z_{it}) & u   & 0   \\
        0       & 0         & r     & s \\
        \end{bmatrix} \\
    W(z_{it}) &= \textrm{logit}^{-1}(b_0 + b_z \, z_{it}) \\
\end{split}
$$ {#eq-aphids}




```{r}
#| label: tbl-aphids
#| tbl-cap: "Model parameters for aphid population dynamics, including the value(s) used in simulations. For variables that change through time, the **Used** column is '$-$'."
tribble(
    ~Parameter,  ~Used,  ~Description ,
    "$n$"    ,  "$4$", "number of aphid stages",
    "$\\mathbf{X}$",  "$-$", "length-$n$ vector of aphid abundances by stage",
    "$g$"    ,  dbl4ltx(pop_info$m, 1), "survival from generalist predators and host-plant death",
    "$\\psi$"    ,  "varies", "survival from *Pseudomonas* bacteria (1 if not present on patch)",
    "$S(\\cdot)$",  "$-$",  "unparasitized aphid density dependence",
    "$z$",  "$-$",  "total alive aphids in a particular patch (i.e., $P + \\sum_{j=1}^{n}{X_{j}}$, where $j$ indexes by aphid stage)",
    "$\\tilde{z}$",  "$-$",  "relative alive aphid abundance in patch $i$ (i.e., $z_{i} / \\sum_{j=1}^{q}{z_{j}}$, where $i$ and $j$ index through space)",
    "$\\mathbf{A}(\\cdot)$",  "$-$",  "length-$n$ vector of the proportion of aphids surviving parasitism",
    "$x$",  "$-$",  "total unparasitized aphids in a particular patch",
    "$Y$",  "$-$",  "total adult, female parasitoids across all patches",
    "$\\mathbf{L}$",  "$**$",  "$n \\times n$ transition matrix that defines transitions among all stages",
    "$K$",  dbl4ltx(pop_info$K, 0),  "controls strength of unparasitized aphid density dependence",
    "$u$",  dbl4ltx(pop_info$surv_j),  "proportion of aphid juveniles that stay as juveniles each day",
    "$s$",  dbl4ltx(pop_info$surv_a),  "proportion of aphid adults that survive each day",
    "$r$",  dbl4ltx(pop_info$recruit),  "proportion of aphid juveniles that transition to adults each day",
    "$f$",  dbl4ltx(pop_info$fecund),  "aphid fecundity",
    "$W(\\cdot)$",  "$-$",  "proportion of winged offspring from non-winged mothers",
    "$b_0$",  dbl4ltx(pop_info$alate_0, 0),  "affects the proportion of winged offspring at low aphid densities",
    "$b_z$",  dbl4ltx(pop_info$alate_1),  "effect of aphid density on winged-aphid production",
) |> 
    kable(escape = FALSE)
```




## Parasitized aphids and mummies


Parasitized aphid ($P$) and mummy ($M$) populations on patch $i$ follow 
the following equations
(see @tbl-parasitized for new parameters):

<!-- -->
$$
\begin{split}
    P_{i,t+1} &= g \, \psi_i \, S_p(z_{it}) \, \left(s_p \, P_{it} + 
        \left[1 - \mathbf{A}(x_{it},\tilde{z}_{it},Y_{t})\right]^{\intercal} \,
            \mathbf{L} \mathbf{X}_{it}
        \right) - r_p \, P_{it} \\
    S_p(z_{it}) &= \left( 1 + \frac{z_{it}}{c \, K} \right)^{-1} \\
    M_{i,t+1} &= g \, \left( M_{it} + r_p \, P_{it} \right) - r_m \, M_{it} \\
\end{split}
$$ {#eq-paras-mummimes}

In this case, the following must be true to avoid negative numbers:
<!-- -->
$$
\begin{split}
    r_m &< g \\
    r_p &< g \, \rho \, S_p(z_{t}) \, s_p \\
\end{split}
\textrm{,}
$$
<!-- -->
\noindent where the minimum value of $S_p(z_{t})$ is $\sim 0.65$ 
even when non-parasitized aphids are at their carrying capacity
($K \, (\lambda - 1)$, where $\lambda$ is the leading 
eigenvalue of $\mathbf{L}$),
which is higher than for parasitized aphids.


```{r}
#| label: tbl-parasitized
#| tbl-cap: "Model parameters for parasitized aphids, mummies, and adult parasitoids, including the value(s) used in simulations. For variables that change through time, the **Used** column is '$-$'. ** indicates that the vector used for $\\mathbf{R}$ was $\\\\{0.1861, 0.0053, 0.1861, 0\\\\}$"
tribble(
    ~Parameter,  ~Used,  ~Description ,
    "$\\mathbf{R}$", "$**$", "length-$n$ vector giving the relative parasitoid attack rates by aphid stage",
    "$a$"    , dbl4ltx(pop_info$a, 2), "overall parasitoid attack rate",
    "$h$"    , dbl4ltx(pop_info$h, 3), "parasitoid attack-rate handling time",
    "$k$"    , dbl4ltx(pop_info$k, 2), "negative binomial distribution's aggregation parameter",
    "$P$"    , "$-$", "parasitized aphid abundance",
    "$S_p(\\cdot)$"  , "$-$", "parasitized aphid density dependence",
    "$c$"  , dbl4ltx(pop_info$K_p_mult, 4), "density dependence for parasitized aphids relative to unparasitized ($0 < c \\le 1$)",
    "$s_p$"  , dbl4ltx(pop_info$s_p), "daily survival for parasitized aphids",
    "$r_p$", dbl4ltx(pop_info$trans_pm), "proportion of parasitized aphids that transition to mummies each day",
    "$M$"  , "$-$", "mummy (parasitized, dead aphid) abundance",
    "$r_m$", dbl4ltx(pop_info$trans_ma), "proportion of mummies that transition to adult parasitoids each day",
    "$s_y$"  , dbl4ltx(pop_info$s_y, 2), "adult parasitoid daily survival",
    "$\\zeta$", "varies", "constant between 0 and 1 that affects the extent to which parasitoids respond to aphid density",
    "$\\gamma(\\cdot)$", "$-$", "number of parasitoids on each patch"
) |> 
    kable(escape = FALSE)
```




## Adult parasitoids and attacks on aphids

There is a single adult parasitoid population across all $q$ patches
that operates as such:

$$
    Y_{t+1} = s_y \, Y_{t} + \frac{1}{2} \, r_m \sum_{i=1}^{q}{M_{it}}
$$ {#eq-adult-paras}


The number of parasitoids on each patch $i$ ($\gamma$) is affected by 
the total number of parasitoids and the
relative number of aphids on that patch ($\tilde{z}$):

$$
    \gamma(\tilde{z}_{it}) = Y_{t} \left[ 
            (1 - \zeta) \, \frac{1}{q} + \zeta \, \tilde{z}_{it}
        \right]
$$ {#eq-wasp-abunds-i}

\noindent where $\zeta$ is a constant between 0 and 1 that affects the extent
to which parasitoids respond to aphid density
([Godfray \& Pacala, 1992](https://doi.org/10.1086/285401)).


Because parasitoid abundances vary across patches, so do the proportion of 
aphids that survive parasitoid attacks
($\mathbf{A}$).
This survival is affected by 
non-parasitized aphid density on the patch $i$ ($x$),
the relative number of aphids on the patch $i$ ($\tilde{z}$), and
adult parasitoid density ($Y$):


$$
\mathbf{A}(x_{it},\tilde{z}_{it},Y_{t}) =  \left[1 + \mathbf{R} \, \frac{a \, \gamma(\tilde{z}_{it}) }{
    k \, (h \, x_{it} + 1)} \right]^{\circ -k}
$$ {#eq-wasp-attacks}

