---
title: "Host-parasitoid model description"
author: "Lucas A. Nell"
execute:
  echo: false
  message: false
  warning: false
format: 
  html:
    html-math-method: katex
vignette: >
  %\VignetteIndexEntry{Host-parasitoid model description}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


```{=html}
<style type="text/css">
  figure.quarto-float-tbl figcaption.quarto-float-caption-top {
    margin-top: .5rem;
    margin-bottom: .25rem;
    text-align:left
  }
</style>
```


```{r}
#| label: setup
#| include: false
library(knitr)
library(kableExtra)
library(gt)
library(gameofclones)
library(aeonia)
# Make a number print in LaTeX friendly format:
dbl4ltx <- function(x, .digits = 4) {
    .fmt <- sprintf("$%%.%if$", .digits)
    x <- sprintf(.fmt, x)
    return(I(x))
}
```


In the equations below, $\odot~$indicates element-wise
multiplication (Hadamard product), 
$^{\circ}~$indicates element-wise power (Hadamard power),
and $^{\intercal}~$indicates matrix transposition.
Vectors and matrices are in bold.


## Aphid population dynamics

Aphid populations ($\mathbf{X}$) on patch $i$ follow the following equations
(see @tbl-aphids for parameters):


<!-- -->
$$
\begin{split}
    \mathbf{X}_i(t+1) &= g \, \rho_i \, S(z_i(t)) \, 
        \mathbf{A}(x_i(t), \hat{z}_i(t), Y(t)) \odot \left[ \mathbf{L} \, \mathbf{X}_i(t) \right] \\
    S(z_i(t)) &= \left( 1 + \frac{z_i(t)}{K} \right)^{-1} \\
    \mathbf{L} &= \begin{pmatrix}
        s_j     & f \, (1{-}W(z_i(t))) & 0     & f   \\
        r       & s_a       & 0     & 0   \\
        0       & f \, W(z_i(t)) & s_j   & 0   \\
        0       & 0         & r     & s_a \\
        \end{pmatrix} \\
    W(z_i(t)) &= \textrm{logit}^{-1}(\delta_0 + \delta_z \, z_i(t)) \\
\end{split}
$$ {#eq-aphids}




```{r}
#| label: tbl-aphids
#| tbl-cap: "Model parameters for aphid population dynamics, including the value(s) used in simulations. For variables that change through time, the **Used** column is '$-$'."
tribble(
    ~Parameter,  ~Used,  ~Description ,
    "$n$"    ,  "$4$", "number of aphid stages",
    "$\\mathbf{X}(t)$",  "$-$", "length-$n$ vector of aphid abundances by stage",
    "$g$"    ,  dbl4ltx(pop_info$m, 1), "survival from generalist predators and host-plant death",
    "$\\rho$"    ,  "varies", "survival from *Pseudomonas* bacteria (1 if not present on patch)",
    "$S(z(t))$",  "$-$",  "unparasitized aphid density dependence",
    "$z(t)$",  "$-$",  "total alive aphids in a particular patch (i.e., $P(t) + \\sum_{j=1}^{n}{X_j(t)}$)",
    "$\\hat{z}(t)$",  "$-$",  "relative alive aphid abundance in patch $i$ (i.e., $z_i(t) / \\sum_{k=1}^{q}{z_k(t)}$)",
    "$\\mathbf{A}$",  "$-$",  "length-$n$ vector of the proportion of aphids surviving parasitism",
    "$x(t)$",  "$-$",  "total unparasitized aphids in a particular patch",
    "$Y(t)$",  "$-$",  "total adult, female parasitoids",
    "$\\mathbf{L}$",  "$**$",  "$n \\times n$ transition matrix that defines transitions among all stages",
    "$K$",  dbl4ltx(pop_info$K, 0),  "controls strength of unparasitized aphid density dependence",
    "$s_j$",  dbl4ltx(pop_info$surv_j),  "proportion of aphid juveniles that stay as juveniles each day",
    "$s_a$",  dbl4ltx(pop_info$surv_a),  "proportion of aphid adults that survive each day",
    "$r$",  dbl4ltx(pop_info$recruit),  "proportion of aphid juveniles that transition to adults each day",
    "$f$",  dbl4ltx(pop_info$fecund),  "aphid fecundity",
    "$W(z(t))$",  "$-$",  "proportion of winged offspring from non-winged mothers",
    "$\\delta_0$",  dbl4ltx(pop_info$alate_0, 0),  "affects the proportion of winged offspring at low aphid densities",
    "$\\delta_z$",  dbl4ltx(pop_info$alate_1),  "effect of aphid density on winged-aphid production",
) |> 
    kable(escape = FALSE)
```




## Parasitized aphids and mummies


Parasitized aphid ($P$) and mummy ($M$) populations on patch $i$ follow 
the following equations
(see @tbl-parasitized for new parameters):

<!-- -->
$$
\begin{split}
    P_i(t+1) &= g \, \rho_i \, S_p(z_i(t)) \, \left(s_p \, P_i(t) + 
        \left[1 - \mathbf{A}(x_i(t),\hat{z}_i(t),Y(t))\right]^{\intercal} \,
            \mathbf{L} \mathbf{X}_i(t)
        \right) - \phi \, P_i(t) \\
    S_p(z_i(t)) &= \left( 1 + \frac{z_i(t)}{\nu \, K} \right)^{-1} \\
    M_i(t+1) &= g \, \left[M_i(t) + \phi \, P_i(t) \right] - \theta \, M_i(t) \\
\end{split}
$$ {#eq-paras-mummimes}

In this case, the following must be true to avoid negative numbers:
<!-- -->
$$
\begin{split}
    \theta &< g \\
    \phi &< g \, \rho \, S_p(z(t)) \, s_p \\
\end{split}
\textrm{,}
$$
<!-- -->
\noindent where the minimum value of $S_p(z(t))$ is $\sim 0.65$ 
even when non-parasitized aphids are at their carrying capacity
($K \, (\lambda - 1)$, where $\lambda$ is the leading 
eigenvalue of $\mathbf{L}$),
which is higher than for parasitized aphids.


```{r}
#| label: tbl-parasitized
#| tbl-cap: "Model parameters for parasitized aphids, mummies, and adult parasitoids, including the value(s) used in simulations. For variables that change through time, the **Used** column is '$-$'. ** indicates that the vector used for $\\mathbf{R}$ was $\\\\{0.1861, 0.0053, 0.1861, 0\\\\}$"
tribble(
    ~Parameter,  ~Used,  ~Description ,
    "$\\mathbf{R}$", "$**$", "length-$n$ vector giving the relative parasitoid attack rates by aphid stage",
    "$a$"    , dbl4ltx(pop_info$a, 2), "overall parasitoid attack rate",
    "$h$"    , dbl4ltx(pop_info$h, 3), "parasitoid attack-rate handling time",
    "$k$"    , dbl4ltx(pop_info$k, 2), "negative binomial distribution's aggregation parameter",
    "$P$"    , "$-$", "parasitized aphid abundance",
    "$S_p(z(t))$"  , "$-$", "parasitized aphid density dependence",
    "$\\nu$"  , dbl4ltx(pop_info$K_p_mult, 4), "density dependence for parasitized aphids relative to unparasitized ($0 < \\nu \\le 1$)",
    "$s_p$"  , dbl4ltx(pop_info$s_p), "daily survival for parasitized aphids",
    "$\\phi$", dbl4ltx(pop_info$trans_pm), "proportion of parasitized aphids that transition to mummies each day",
    "$M$"  , "$-$", "mummy (parasitized, dead aphid) abundance",
    "$\\theta$", dbl4ltx(pop_info$trans_ma), "proportion of mummies that transition to adult parasitoids each day",
    "$s_y$"  , dbl4ltx(pop_info$s_y, 2), "adult parasitoid daily survival",
    "$\\zeta$", "varies", "constant between 0 and 1 that affects the extent to which parasitoids respond to aphid density"
) |> 
    kable(escape = FALSE)
```




## Adult parasitoids and attacks on aphids

There is a single adult parasitoid population across all $q$ patches
that operates as such:

$$
    Y(t+1) = s_y \, Y(t) + \frac{1}{2} \, \theta \sum_{i=1}^{q}{M_i(t)}
$$ {#eq-adult-paras}


The number of parasitoids on each patch $i$ ($\gamma$) is affected by 
the total number of parasitoids and the
relative number of aphids on that patch ($\hat{z}_i$):

$$
    \gamma(\hat{z}_i(t)) = Y(t) \left[ 
            (1 - \zeta) \, \frac{1}{q} + \zeta \, \hat{z}_i(t)
        \right]
$$ {#eq-wasp-abunds-i}

\noindent where $\zeta$ is a constant between 0 and 1 that affects the extent
to which parasitoids respond to aphid density
([Godfray \& Pacala, 1992](https://doi.org/10.1086/285401)).


Because parasitoid abundances vary across patches, so do the proportion of 
aphids that survive parasitoid attacks
($\mathbf{A}$).
This survival is affected by 
non-parasitized aphid density on the patch $i$ ($x$),
the relative number of aphids on the patch $i$ ($\hat{z}_i$), and
adult parasitoid density ($Y$):


$$
\mathbf{A}(x(t),\hat{z}(t),Y(t)) =  \left[1 + \mathbf{R} \, \frac{a \, \gamma(\hat{z}_i(t)) }{
    k \, (h \, x(t) + 1)} \right]^{\circ -k}
$$ {#eq-wasp-attacks}

