---
title: "Model description"
author: "Lucas A. Nell"
execute:
  echo: false
  message: false
  warning: false
format: 
  html:
    toc: true
    toc-depth: 4
    html-math-method: katex
vignette: >
  %\VignetteIndexEntry{Model description}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


```{=html}
<style type="text/css">
  figure.quarto-float-tbl figcaption.quarto-float-caption-top {
    margin-top: .5rem;
    margin-bottom: .25rem;
    text-align:left
  }
</style>
```


```{r}
#| label: setup
#| include: false
library(knitr)
library(kableExtra)
library(gt)
library(gameofclones)
library(aeonia)
# Make a number print in LaTeX friendly format:
dbl4ltx <- function(x, .digits = 4) {
    .fmt <- sprintf("$%%.%if$", .digits)
    x <- sprintf(.fmt, x)
    return(I(x))
}
```


In the equations below, $\odot~$indicates element-wise
multiplication (Hadamard product), 
$^{\circ}~$indicates element-wise power (Hadamard power),
and $^{\intercal}~$indicates matrix transposition.
Vectors and matrices are in bold.
Unless otherwise noted, indexing occurs first by space, then by time,
so $x_{it}$ is for patch $i$ at time $t$.


# Host--parasitoid model

## Aphid population dynamics

Aphid populations ($\mathbf{X}$) on patch $i$ follow the following equations
(see @tbl-aphids for parameters):


<!-- -->
$$
\begin{split}
    \mathbf{X}_{i,t+1} &= g \, \psi_i \, S(z_{it}) \, 
        \mathbf{A}(x_{it}, \tilde{z}_{it}, Y_{t}) \odot \left[ \mathbf{L} \, \mathbf{X}_{it} \right] \\
    S(z_{it}) &= \left( 1 + \frac{z_{it}}{K} \right)^{-1} \\
    \mathbf{L} &= \begin{bmatrix}
        u     & f \, (1{-}W(z_{it})) & 0     & f   \\
        r       & s       & 0     & 0   \\
        0       & f \, W(z_{it}) & u   & 0   \\
        0       & 0         & r     & s \\
        \end{bmatrix} \\
    W(z_{it}) &= \left(1 + 10^{ (b_{\textrm{infl.}} - z_{it}) \, b_{\textrm{slope}} } \right)^{-1} \\
\end{split}
$$ {#eq-aphids}




```{r}
#| label: tbl-aphids
#| tbl-cap: "Model parameters for aphid population dynamics, including the value(s) used in simulations. For variables that change through time, the **Used** column is '$-$'."
tribble(
    ~Parameter,  ~Used,  ~Description ,
    "$n$"    ,  "$4$", "number of aphid stages",
    "$\\mathbf{X}$",  "$-$", "length-$n$ vector of aphid abundances by stage",
    "$g$"    ,  dbl4ltx(pop_info$m, 1), "survival from generalist predators and host-plant death",
    "$\\psi$"    ,  "varies", "survival from *Pseudomonas* bacteria (1 if not present on patch)",
    "$S(\\cdot)$",  "$-$",  "unparasitized aphid density dependence",
    "$z$",  "$-$",  "total alive aphids in a particular patch (i.e., $P + \\sum_{j=1}^{n}{X_{j}}$, where $j$ indexes by aphid stage)",
    "$\\tilde{z}$",  "$-$",  "relative alive aphid abundance in patch $i$ (i.e., $z_{i} / \\sum_{j=1}^{q}{z_{j}}$, where $i$ and $j$ index through space)",
    "$\\mathbf{A}(\\cdot)$",  "$-$",  "length-$n$ vector of the proportion of aphids surviving parasitism",
    "$x$",  "$-$",  "total unparasitized aphids in a particular patch",
    "$Y$",  "$-$",  "total adult, female parasitoids across all patches",
    "$\\mathbf{L}$",  "$**$",  "$n \\times n$ transition matrix that defines transitions among all stages",
    "$K$",  dbl4ltx(pop_info$K, 0),  "controls strength of unparasitized aphid density dependence",
    "$u$",  dbl4ltx(pop_info$surv_j),  "proportion of aphid juveniles that stay as juveniles each day",
    "$s$",  dbl4ltx(pop_info$surv_a),  "proportion of aphid adults that survive each day",
    "$r$",  dbl4ltx(pop_info$recruit),  "proportion of aphid juveniles that transition to adults each day",
    "$f$",  dbl4ltx(pop_info$fecund),  "aphid fecundity",
    "$W(\\cdot)$",  "$-$",  "proportion of winged offspring from non-winged mothers",
    "$b_{\\textrm{infl.}}$",  dbl4ltx(pop_info$alate_infl, 0),  "inflection point for the sigmoid relationship between aphid density and alate offspring proportion",
    "$b_{\\textrm{slope}}$",  dbl4ltx(pop_info$alate_slope, 5),  "slope for the sigmoid relationship between aphid density and alate offspring proportion",
) |> 
    kable(escape = FALSE)
```




## Parasitized aphids and mummies


Parasitized aphid ($P$) and mummy ($M$) populations on patch $i$ follow 
the following equations
(see @tbl-parasitized for new parameters):

<!-- -->
$$
\begin{split}
    P_{i,t+1} &= g \, \psi_i \, S_p(z_{it}) \, \left(s_p \, P_{it} + 
        \left[1 - \mathbf{A}(x_{it},\tilde{z}_{it},Y_{t})\right]^{\intercal} \,
            \mathbf{L} \mathbf{X}_{it}
        \right) - r_p \, P_{it} \\
    S_p(z_{it}) &= \left( 1 + \frac{z_{it}}{c \, K} \right)^{-1} \\
    M_{i,t+1} &= g \, \left( M_{it} + r_p \, P_{it} \right) - r_m \, M_{it} \\
\end{split}
$$ {#eq-paras-mummimes}

In this case, the following must be true to avoid negative numbers:
<!-- -->
$$
\begin{split}
    r_m &< g \\
    r_p &< g \, \psi \, S_p(z_{t}) \, s_p \\
\end{split}
\textrm{,}
$$
<!-- -->
\noindent where the minimum value of $S_p(z_{t})$ is $\sim 0.65$ 
even when non-parasitized aphids are at their carrying capacity
($K \, (\lambda - 1)$, where $\lambda$ is the leading 
eigenvalue of $\mathbf{L}$),
which is higher than for parasitized aphids.


```{r}
#| label: tbl-parasitized
#| tbl-cap: "Model parameters for parasitized aphids, mummies, and adult parasitoids, including the value(s) used in simulations. For variables that change through time, the **Used** column is '$-$'. ** indicates that the vector used for $\\mathbf{R}$ was $\\\\{0.1861, 0.0053, 0.1861, 0\\\\}$"
tribble(
    ~Parameter,  ~Used,  ~Description ,
    "$\\mathbf{R}$", "$**$", "length-$n$ vector giving the relative parasitoid attack rates by aphid stage",
    "$a$"    , dbl4ltx(pop_info$a, 2), "overall parasitoid attack rate",
    "$h$"    , dbl4ltx(pop_info$h, 3), "parasitoid attack-rate handling time",
    "$k$"    , dbl4ltx(pop_info$k, 2), "negative binomial distribution's aggregation parameter",
    "$P$"    , "$-$", "parasitized aphid abundance",
    "$S_p(\\cdot)$"  , "$-$", "parasitized aphid density dependence",
    "$c$"  , dbl4ltx(pop_info$K_p_mult, 4), "density dependence for parasitized aphids relative to unparasitized ($0 < c \\le 1$)",
    "$s_p$"  , dbl4ltx(pop_info$s_p), "daily survival for parasitized aphids",
    "$r_p$", dbl4ltx(pop_info$trans_pm), "proportion of parasitized aphids that transition to mummies each day",
    "$M$"  , "$-$", "mummy (parasitized, dead aphid) abundance",
    "$r_m$", dbl4ltx(pop_info$trans_ma), "proportion of mummies that transition to adult parasitoids each day",
    "$s_y$"  , dbl4ltx(pop_info$s_y, 2), "adult parasitoid daily survival",
    "$\\zeta$", "varies", "constant between 0 and 1 that affects the extent to which parasitoids respond to aphid density",
    "$\\gamma(\\cdot)$", "$-$", "number of parasitoids on each patch"
) |> 
    kable(escape = FALSE)
```




## Adult parasitoids and attacks on aphids

There is a single adult parasitoid population across all $q$ patches
that operates as such:

$$
    Y_{t+1} = s_y \, Y_{t} + \frac{1}{2} \, r_m \sum_{i=1}^{q}{M_{it}}
$$ {#eq-adult-paras}


The number of parasitoids on each patch $i$ ($\gamma$) is affected by 
the total number of parasitoids and the
relative number of aphids on that patch ($\tilde{z}$):

$$
    \gamma(\tilde{z}_{it}) = Y_{t} \left[ 
            (1 - \zeta) \, \frac{1}{q} + \zeta \, \tilde{z}_{it}
        \right]
$$ {#eq-wasp-abunds-i}

\noindent where $\zeta$ is a constant between 0 and 1 that affects the extent
to which parasitoids respond to aphid density
([Godfray \& Pacala, 1992](https://doi.org/10.1086/285401)).


Because parasitoid abundances vary across patches, so do the proportion of 
aphids that survive parasitoid attacks
($\mathbf{A}$).
This survival is affected by 
non-parasitized aphid density on the patch $i$ ($x$),
the relative number of aphids on the patch $i$ ($\tilde{z}$), and
adult parasitoid density ($Y$):


$$
\mathbf{A}(x_{it},\tilde{z}_{it},Y_{t}) =  \left[1 + \mathbf{R} \, \frac{a \, \gamma(\tilde{z}_{it}) }{
    k \, (h \, x_{it} + 1)} \right]^{\circ -k}
$$ {#eq-wasp-attacks}





<!---
================================================================================
================================================================================
-->
# Spatial and disease model

Patches are organized as a two-dimensional, $n_x \times n_y$ landscape.
Each patch either contains *Pseudomonas* ($\psi < 1$) or not ($\psi = 1$),
and *Pseudomonas* does not move once the simulations start.
The patch in the upper-left corner of the landscape ($x = 1$, $y = 1$)
starts as the only patch with plants infectious with the virus, and 
the virus is only transmitted between patches through
the dispersal of alates (winged aphids).
We assume that all virus transmission is via nonpersistent transmission (NPT)
where the virus does not infect the aphid, but rather loosely clings
to aphid mouthparts.
We simulate NPT since this is how most aphid-transmitted viruses are spread
([Gray \& Banerjee 1999](https://doi.org/10.1128/mmbr.63.1.128-148.1999),
[Nault 1997](https://doi.org/10.1093/aesa/90.5.521)).



## Alate dispersal

Every day, the number of adult alates that leaves their starting patch
($n_{\textrm{fly}}$) is sampled as such:

$$
n_{\textrm{fly}} \sim \textrm{Binomial}(\lfloor X_4 \rfloor, p_{\textrm{fly}})
$$

\noindent where
$\lfloor \cdot \rfloor$ is the floor function,
$X_4$ is the density of adult alates,
and 
$p_{\textrm{fly}}$ is the probability that an adult alate leaves their plant
each day.

Alate flight paths are simulated individually.
Each alate that leaves a patch can travel to any other patch that is
within alates' dispersal radius ($r_{\textrm{fly}}$).
The probability that the alate will choose patch $j$ from patch $i$ is given by

$$
\textrm{Pr}(i \rightarrow j) = \frac{ \omega_j }{ 
    \sum\limits_{k \in \mathbf{D}_i}{ \omega_k } }
$$ {#eq-alate-samp-prob}

\noindent where $\omega_j$ is the sampling weight of plant $j$,
and
$\mathbf{D}_i$ is a vector of indices for all plants within 
$r_{\textrm{fly}}$ of plant $i$.
Sampling weights depend on the state of the plants in the patch:

$$
\omega_i = \begin{cases}
    \nu & \textrm{if infectious with virus} \\
    \rho^{-1} & \textrm{if } \textit{Pseudomonas} \textrm{ present} \\
    \nu / \rho & \textrm{if both} \\
    1 & \textrm{if none} \\
\end{cases}
$$ {#eq-sampling-weights}

\noindent where 
$\nu$ is the attractant effect of the virus ($\nu > 1$),
and
$\rho$ is the repellent effect of *Pseudomonas* ($\rho > 1$).

Once an alate flies to a new patch, it will stay at the new patch with
probability $w$. It will continue to fly to new patches until it chooses
to stay indefinitely.


## Virus spread

We assume that alates probe plants on each visit, and if alates
visit a patch containing plants infectious with the virus (including the patch 
they start on), the virus loads onto the alates with probability $\delta_a$.
If an alate loaded with the virus probes a patch of uninfected plants,
the probability that the patch is exposed to the virus is $\delta_p$.
We assume that once an alate is loaded with the virus, it can deliver the
virus to all subsequent patches it probes on that day.
A patch that is exposed to the virus is not immediately able to pass the virus
onto other alates.
Patches transition from exposed to infectious after 7 days, which is about
how long it takes in the cucumber mosaic virus
([Davis & Hampton, 1986](https://www.apsnet.org/publications/phytopathology/backissues/Documents/1986Articles/Phyto76n10_999.PDF)),
which spreads via NPT by aphids.

